Evaluating Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe using radiocarbon data

### 2019-02-05 article/03_Results.Rmd
more work on the description of results

Figure \ref{development_burial_type} and \ref{development_burial_construction} show the region-wise time series of trait occurrence. If we ignore all other archaeological knowledge, we can construct the following narrative from them: \newline

In the course of the Bronze Age in Europe there was a trend away from body burial towards cremation. Around 2200 calBC cremation burials were almost unknown in Poland, Southern Germany, Northeastern France and Northern Germany. In the Northwest, in Great Britain and the Benelux area, and in the Southeast, in Austria and the Czech Republic, however, there were early cremation traditions. These contexts could be discussed as areas of origin of the later omnipresent phenomenon. While cremation burials lost importance in Austria and the Czech Republic as well as in the Benelux region during the Early Bronze Age and only regained relevance there in the Middle Bronze Age, their share increased steadily in England. Inhumation and cremation were balanced here for a long time. The situation was similar in Southern Scandinavia, where the proportion of cremation burials was much lower until the Late Bronze Age. In Poland and northern Germany, cremation became relatively suddenly and completely established with the beginning of the Late Bronze Age. In Northeastern France and Northern Germany this change took place already in the Middle Bronze Age. In both contexts inhumation continued to play an important role.

@@

This narrative is wrong or inaccurate in some aspects. Please see the analysis in section \ref{time-series-discussion}.

@@

Table \ref{tab:distanceandcorrelationtable*} shows that the distance networks of *burial type* and *burial construction* are not equivalent. There is also no significant negative correlation. The degree of agreement between two regions in the question of inhumation and cremation is therefore not suitable for predicting the similarity of developments in these regions concerning the dominance of flat or mound graves. Only for the time slot at the transition to the Middle Bronze Age from 1800 to 1600 calBC the 5% significance level of the Mantle test is almost undercut, thus suggesting a statistically significant relationship.

For the relationship between the SED network based on *burial type* and the spatial distance several observations result from Table \ref{tab:distanceandcorrelationtable*}: Cultural distance calculated from these to variables does not seem to correlate with spatial distance. A significant correlation between the distance measures only reveals itself for the Early Bronze Age time slot 2000 to 1800 calBC: Only in the Early Bronze Age, long before the universal switch to the Urnfield burial tradition, large regions directly adjacent to each other tend to undergo the same developments rather than the ones further away from each other. In the Middle Bronze Age the SED developments of the *burial type* variable undergo the peaks already observed in Figure \ref{sed_region_matrix}. The correlation between cultural and spatial distance over these peaks is low and even tilts between 1400 and 1200 calBC, at the transition to the Late Bronze Age, into the negative -- although not significantly. The cultural distance between several geographically distant regions is then smaller than between several directly adjacent regions. Looking at the SED network of *burial construction* (Table \ref{tab:distanceandcorrelationtable*}), similar conclusions can be drawn as in the *burial type* context: The distance values here are larger overall, and the SED differences within the individual spatial distance classes are clearly more accentuated. As well as in *burial type* the Early Bronze Age is by far the phase with the most substantial spatial correlation, though it is only significant for the time slot between 1800 and 1400 calBC. There is no time period with negative mean correlation for *burial construction* cultural distance and spatial distance, but also here the correlation coefficient approaches zero in the Late Bronze Age.

The results of the simulation as presented in Figure \ref{simulation} confirm the expectation: In case of equal interaction intensity $I$ of separate populations, the correlation of cultural and spatial distance is weak (Figure \ref{simulation} A vs. C vs. E). It appears to be normally distributed around zero for many simulation runs, although individual runs can still yield remarkably high positive and negative correlation for certain time slots. The fluctuation of the correlation within one run can be strong and seems to increase for bigger values of $m_i$. If the interaction matrix $I$ is predefined to represent the spatial distance matrix (Figure \ref{simulation} BDF), then the mean correlation per time slot increases. The correlation increases as well with the degree of interaction $m_i$ (Figure \ref{simulation} D vs. F). Counter-intuitively the fluctuation of the correlation is also very high here (Figure \ref{simulation} E vs. F). The correlation development observed for the variables *burial type* and *burial construction* fits to simulation runs in all presets, but becomes less likely if we assume $I$ to represent spatial distance and a high degree of interaction $m_i$ (Figure \ref{simulation} A vs. F). That's mostly because of the relatively strong negative correlation of cultural and spatial distance in the Middle and Late Bronze Age (1400 to 1000 calBC).

>>>>

Figure \ref{development_burial_type} and \ref{development_burial_construction} show the region-wise time series of trait occurrence. If we consider these results in isolation, we can construct the following narrative from them: \newline

The overall picture is that during the Bronze Age in Europe there was a trend away from body burial towards cremation. Around 2200 calBC cremation burials were almost unknown in Poland, Southern Germany, Northeastern France and Northern Germany. In the Northwest, in Great Britain and the Benelux area, and in the Southeast, in Austria and the Czech Republic, however, there were early cremation traditions. These contexts could be discussed as areas of origin of the later omnipresent phenomenon. While cremation burials lost importance in Austria and the Czech Republic as well as in the Benelux region during the Early Bronze Age and only regained relevance there in the Middle Bronze Age, their share increased steadily in England. Inhumation and cremation were balanced here for a long time. The situation was similar in Southern Scandinavia, where the proportion of cremation burials was much lower until the Late Bronze Age. In Poland and northern Germany, cremation became relatively suddenly and completely established with the beginning of the Late Bronze Age. In Northeastern France and Northern Germany this change took place already in the Middle Bronze Age. In both contexts inhumation continued to play an important role.

@@

@@

Table \ref{tab:distanceandcorrelationtable*} C shows that the distance networks of *burial type* and *burial construction* are not equivalent. There is also no significant negative correlation. The degree of agreement between two regions in the question of inhumation and cremation is therefore not suitable for predicting the similarity of developments in these regions concerning the dominance of flat or mound graves. Only for the time slot at the transition to the Middle Bronze Age from 1800 to 1600 calBC the 5% significance level of the Mantle test is almost undercut, thus suggesting a statistically significant relationship.

For the relationship between the SED networks and the spatial distance network several observations result from Table \ref{tab:distanceandcorrelationtable*} AB: Cultural distance generally does not seem to correlate with spatial distance. For the variable *burial type* an exception with significant correlation between the distance measures only reveals itself for the Early Bronze Age time slot 2000 to 1800 calBC. In the Early Bronze Age, long before the universal switch to the Urnfield burial tradition, large regions directly adjacent to each other tend to see the same developments. In the Middle Bronze Age the SED developments of the *burial type* variable undergo the peaks already observed. The correlation between cultural and spatial distance during these peaks is low and even tilts between 1400 and 1200 calBC, at the transition to the Late Bronze Age, into the negative -- although not significantly. The cultural distance between several geographically distant regions is then smaller than between several directly adjacent regions. Looking at the SED network of *burial construction*, similar conclusions can be drawn: The distance values here are larger overall, and the SED differences within the individual spatial distance classes are clearly more accentuated. As well as in *burial type* the Early Bronze Age is by far the phase with the most substantial spatial correlation, though it is only significant for the time slot between 1800 and 1400 calBC. There is no time period with negative mean correlation for *burial construction* cultural distance and spatial distance, but also here the correlation coefficient approaches zero in the Late Bronze Age.

The simulation as presented in Figure \ref{simulation} was implemented to contextualize this observation of low correlation between spatial and cultural distance. Some general remarks: In case of equal interaction intensity $I$ of separate populations, the correlation of cultural and spatial distance is weak (Figure \ref{simulation} A vs. C vs. E). It appears to be normally distributed around zero for many simulation runs, although individual runs can still yield high positive and negative correlation for certain time slots. The fluctuation of the correlation within one run can be strong and seems to increase for greater values of $m_i$. If the interaction matrix $I$ is predefined to represent the spatial distance matrix (Figure \ref{simulation} BDF), then the mean correlation per time slot increases. The correlation increases as well with the degree of interaction $m_i$ (Figure \ref{simulation} D vs. F). Counter-intuitively the fluctuation of the correlation is also very high here (Figure \ref{simulation} E vs. F). The correlation development observed for the variables *burial type* and *burial construction* fits to simulation runs in all presets, but becomes less likely if we assume $I$ to represent spatial distance and a high degree of interaction $m_i$ (Figure \ref{simulation} A vs. F). That's mostly because of the relatively strong negative correlation of cultural and spatial distance in the Middle and Late Bronze Age (1400 to 1000 calBC). Despite this tendency, the real-world correlation time series are still within the margins of potential toy model events. Therefore it is not possible to rule out the influence of spatial proximity as an important factor for the Urnfield burial rite diffusion. On the contrary, the simulation results challenge an overly simplistic interpretation of Table \ref{tab:distanceandcorrelationtable*}.
